---
title: "Making and publishing R package repository snapshots"
output: html_document
vignette: >
  %\VignetteIndexEntry{Making and publishing package repository snapshots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


An R package repository is simply directory containing the R package files, accompanied a file named `PACKAGES` that stores and index and metadata needed by the R package manager. Optionally there are also `PACKAGES.gz` and `PACKAGES.rds` files, storing exactly the same data in gzipped or rds format respectively.

R package repositories are static files and can be hosted on any public HTTP server. The folders need to be in the following directory structure for the R package manager to find them:

```
./src/contrib/                        ← dir with source packages
./bin/windows/contrib/{x,y}/          ← dir with windows binaries for R major version x,y
./bin/macosx/contrib/{x,y}/           ← dir with macos binaries for R major version x,y
```

These paths are relative to the `repos` parameter in `install.packages()`.



## Snapshotting a full package repository

Use `repo_snapshot()` to download an entire R package repository:

```{r}
library(snapshot)
destdir <- 'r-rust-snapshot'
db <- repo_snapshot('https://r-rust.r-universe.dev', destdir)

# See what we got
list.files(destdir, recursive = TRUE, full.names = TRUE)
```

By default `repo_snapshot()` only downloads the source packages. Usually R can install packages from source, however for packages with compiled C/C++/Rust/Fortran code, it may convenient for users to also provide the pre-compiled binary packages:


```{r}
db <- repo_snapshot('https://r-rust.r-universe.dev', 'r-rust-snapshot',
                    win_binaries = TRUE, mac_binaries = TRUE, bin_versions = "4.2")
# See what we got
list.files(destdir, recursive = TRUE, full.names = TRUE)
```

Note that binary packages for Windows and MacOS can only be used with exactly that R version. You can pass a vector to `bin_versions` in case you want to download binaries for multiple versions of R (if the server has those available).

## Create a custom snapshot

If an R package repository is large, and you are only interested some of the packages, you can create a custom snapshot of a subset of the repository by specifying the `packages` argument:


```{r}
db <- repo_snapshot('https://ropensci.r-universe.dev', 'my-snapshot', packages = c('magick', 'tesseract', 'unrtf'))
list.files('my-snapshot', recursive = TRUE, full.names = TRUE)
```

This automatically includes the proper `PACKAGES` index files for the custom snapshot subset:

```{r}
db <- read.dcf("my-snapshot/src/contrib/PACKAGES")
db[,c("Package", "Version", "License")]
```


## Installing packages from a local snapshot (offline)

Usually R package repositories are hosted via the web, but it is also possible to use a local package repository (offline). The way to do this is to set a `repos` parameter in R to the full path of the snapshot directory, prefixed with `"file://"`. For example, suppose we want to install packages from the snapshot we just generated:

```{r}
(repo_local <- paste0("file://", normalizePath('my-snapshot')))

# Show available packages in the snapshot
available.packages(repos = repo_local)[,c("Package", "Version", "License")]

# Install a package from the local snapshot
install.packages('unrtf', repos = repo_local, type = 'source')
```



## Hosting snapshots on GitHub

## Archiving snapshots in a container registry


```{r}
unlink('r-rust-snapshot', recursive = TRUE)

```
